---
title: "Introduction to HiContacts"
author: "Jacques Serizay"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{Introduction to HiContacts}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, eval = TRUE, echo=FALSE, results="hide", warning=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>", 
    crop = NULL
)
suppressPackageStartupMessages({
    library(ggplot2)
    library(dplyr)
    library(GenomicRanges)
    library(HiCExperiment)
    library(HiContactsData)
    library(HiContacts)
})
```

# Citing HiContacts

```{r}
citation('HiContacts')
```

# Basics: importing `.(m)/cool` files as `HiCExperiment` objects

The `HiCExperiment` package provides classes and methods to import an .(m)cool 
file in R. The `HiContactsData` package gives access to a range of toy 
datasets stored by Bioconductor in the `ExperimentHub`. 

```{r}
library(HiCExperiment)
library(HiContacts)
cool_file <- HiContactsData::HiContactsData('yeast_wt', format = 'cool')
hic <- HiCExperiment::import(cool_file, format = 'cool')
hic
```

# Plotting matrices 

## Plot matrix heatmaps

The `plotMatrix` function takes a `HiCExperiment` object and plots it as a heatmap.  
Use the `use.scores` argument to specify which type of interaction scores to use 
in the contact maps (e.g. `raw`, `balanced`, ...). By default, `plotMatrix()` 
looks for balanced scores. If they are not stored in the original `.(m)/cool` file, 
`plotMatrix()` simply takes the first scores available. 

```{r}
plotMatrix(hic, use.scores = 'balanced', limits = c(-4, -1), dpi = 120)
```

## Plot loops

Loops can be plotted on top of Hi-C matrices by providing a `GInteractions` 
object to the `loops` argument. 

*Note:* 
Loops in `.bedpe` format can be imported in R using the `import()` function, and 
converted into `GInteractions` with the 
`InteractionSet::makeGInteractionsFromGRangesPairs()` function. 

```{r}
mcool_file <- HiContactsData::HiContactsData('yeast_wt', format = 'mcool')
loops <- system.file("extdata", 'S288C-loops.bedpe', package = 'HiContacts') |> 
    BiocIO::import() |> 
    InteractionSet::makeGInteractionsFromGRangesPairs()
p <- HiCExperiment::import(mcool_file, format = 'mcool', focus = 'IV') |> 
    plotMatrix(loops = loops, limits = c(-4, -1), dpi = 120)
```

## Plot borders

```{r}
borders <- system.file("extdata", 'S288C-borders.bed', package = 'HiContacts') |> 
    BiocIO::import()
p <- HiCExperiment::import(mcool_file, format = 'mcool', focus = 'IV') |> 
    plotMatrix(loops = loops, borders = borders, limits = c(-4, -1), dpi = 120)
```

# Arithmetics

## Subsetting a contact map 

```{r}
hic <- HiCExperiment::import(mcool_file, format = 'mcool', focus = "II", resolution = 2000)
hic_subset <- hic['II:20000-50000']
InteractionSet::boundingBox(interactions(hic))
InteractionSet::boundingBox(interactions(hic_subset))
```

## Computing autocorrelated contact map 

```{r}
mcool_file <- HiContactsData('mESCs', format = 'mcool')
hic <- import(mcool_file, format = 'mcool', focus = 'chr2', resolution = 160000)
hic <- autocorrelate(hic)
scores(hic)
summary(scores(hic, 'autocorrelated'))
```

## Detrending contact map (map over expected)

```{r}
hic <- import(mcool_file, format = 'mcool', focus = 'chr18:20000000-35000000', resolution = 40000)
detrended_hic <- detrend(hic)
cowplot::plot_grid(
    plotMatrix(detrended_hic, use.scores = 'expected', dpi = 120),
    plotMatrix(detrended_hic, use.scores = 'detrended', scale = 'linear', limits = c(-3, 3), dpi = 120)
)
```

## Summing two maps

```{r}
mcool_file_1 <- HiContactsData('yeast_eco1', format = 'mcool')
mcool_file_2 <- HiContactsData('yeast_wt', format = 'mcool')
hic_1 <- import(mcool_file_1, format = 'mcool', focus = 'II', resolution = 2000)
hic_1 <- hic_1['II:1-300000']
hic_2 <- import(mcool_file_2, format = 'mcool', focus = 'II', resolution = 2000)
hic_2 <- hic_2['II:1-300000']
merged_hic <- merge(hic_1, hic_2) 
hic_1
hic_2
merged_hic
```

## Computing ratio between two maps

```{r}
mcool_file_1 <- HiContactsData('yeast_eco1', format = 'mcool')
mcool_file_2 <- HiContactsData('yeast_wt', format = 'mcool')
hic_1 <- import(mcool_file_1, format = 'mcool', focus = 'II', resolution = 2000)
hic_2 <- import(mcool_file_2, format = 'mcool', focus = 'II', resolution = 2000)
div_hic <- divide(hic_1, by = hic_2) 
div_hic
p <- cowplot::plot_grid(
    plotMatrix(hic_1, use.scores = 'balanced', scale = 'log10', limits = c(-4, -1)),
    plotMatrix(hic_2, use.scores = 'balanced', scale = 'log10', limits = c(-4, -1)),
    plotMatrix(div_hic, use.scores = 'ratio', scale = 'log2', limits = c(-2, 2), cmap = bwrColors())
)
```

# Contact map analysis

## Virtual 4C

```{r}
mcool_file <- HiContactsData('mESCs', format = 'mcool')
hic <- import(mcool_file, format = 'mcool', focus = 'chr18:20000000-35000000', resolution = 40000)
v4C <- virtual4C(hic, viewpoint = GRanges('chr18:31000000-31050000'))
plot4C(v4C, aes(x = center, y = score))
```

## Cis-trans ratios

```{r}
mcool_file <- HiContactsData('yeast_wt', format = 'mcool')
hic <- import(mcool_file, format = 'mcool', resolution = 1000)
cisTransRatio(hic)
```

## P(s)

```{r}
# Without a pairs file
mcool_file <- HiContactsData('yeast_wt', format = 'mcool')
hic <- import(mcool_file, format = 'mcool', resolution = 1000)
ps <- distanceLaw(hic)
plotPs(ps, aes(x = binned_distance, y = norm_p))

# With a pairs file
pairsFile(hic) <- HiContactsData('yeast_wt', format = 'pairs.gz')
ps <- distanceLaw(hic)
plotPs(ps, aes(x = binned_distance, y = norm_p))
plotPsSlope(ps, aes(x = binned_distance, y = slope))

# Comparing P(s) curves
c1 <- import(
    HiContactsData('yeast_wt', format = 'mcool'), 
    format = 'mcool',
    resolution = 1000, 
    pairsFile = HiContactsData('yeast_wt', format = 'pairs.gz')
)
c2 <- import(
    HiContactsData('yeast_eco1', format = 'mcool'), 
    format = 'mcool',
    resolution = 1000, 
    pairsFile = HiContactsData('yeast_eco1', format = 'pairs.gz')
)
ps_1 <- distanceLaw(c1) |> mutate(sample = 'WT')
ps_2 <- distanceLaw(c2) |> mutate(sample = 'Eco1-AID')
ps <- rbind(ps_1, ps_2)
plotPs(ps, aes(x = binned_distance, y = norm_p, group = sample, color = sample))
plotPsSlope(ps, aes(x = binned_distance, y = slope, group = sample, color = sample))
```

# Session info

```{r}
sessionInfo()
```
